---
title: "blastn_tax_analysis"
author: "Kimberly Ledger"
date: "2023-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries}
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
```

```{r load-output-from-taxonkit}
taxonomy <-read.delim("/genetics/edna/workdir/crabs/20230216/trimmed/filtered/outputs/blastn_taxlineage.txt", header = FALSE, na.strings=c(""," ","NA"))

head(taxonomy)
```


# clean up the header a bit
```{r}
# use the full taxonomy rather than the seq id to collapse identical entries
tax_df <- taxonomy %>%
  filter(V4 > 100) %>% # make sure all retained matches are >100 bp
  select(-V2, -V5, -V6, -V7, -V14, -V8, -V9, -V10, -V11, -V12, -V15, -V16) %>%  #remove unnecessary columns
  group_by(V1, V17) %>% # group by the sequence key and the full taxonomy to reduce duplicate entries
  unique() %>% # doing that reduced the number of entries from 146k to 17k
  rename(qseqid=V1, perc_id=V3, length=V4, sscinames=V13, taxonomy=V17) %>% #rename headers
  filter(!str_detect(sscinames, "environmental")) %>% # filter out any environmental samples
  filter(!str_detect(sscinames, "synthetic")) #%>% # filter out any synthetic "samples"
  #filter(perc_id >= 98) # seems like some of the matches below 98% are dubious (jellyfish and herring <1% different??)
```

```{r how-many-total-ASVs-w-taxonomy}
tax_df %>% 
  ungroup() %>%
  select(qseqid) %>%
  unique() %>%
  nrow
```

# reformat table 
```{r}
# formatting the taxonomy variables
taxon_df <- tax_df %>%
  filter(str_detect(taxonomy, ";")) %>%
  separate(taxonomy, into=c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = ";")
```


## Sort and clean-up the results based on % identity

```{r}
# find species that include numbers or "sp."
taxon_df %>%
  filter(str_detect(sscinames, " x "))
```

only hybrid seq is C.japonicus x C. opilio... will just remove these ID's since other Chionoecetes ID's are present for the same ASVs
```{r}
taxon_df <- taxon_df %>%
  filter(species != "Chionoecetes japonicus x Chionoecetes opilio")
```

```{r}
# 1. sequences that are unambiguously a single species
single_spp_seqs <- taxon_df %>% 
  group_by(qseqid) %>%
  add_tally(name = "n_taxa") %>%
  filter(n_taxa == 1)
```
three asv's are single-species

```{r}
# remove the single-species seqs from the dataframe and then rank the hits by % identity for the remaining seqs
seq_id_diff <- taxon_df %>%
  anti_join(., single_spp_seqs) %>%
  select(-length) %>%
  group_by(qseqid, species, genus, family, order, class, phylum, kingdom) %>%
    mutate(seq_percID = max(perc_id)) %>%
    group_by(qseqid, species, genus, family, order, class, phylum, kingdom, seq_percID) %>%
  summarise(max(seq_percID)) %>% # take just the top hit for each taxon (for each sequence)
  select(-`max(seq_percID)`) %>%
  ungroup() %>%
  group_by(qseqid) %>%
      mutate(id_rank = rank(desc(seq_percID), ties.method = "min")) %>% # rank the taxonomic hits per sequence by % id
       mutate(top_perc = max(seq_percID)) %>% # designate the highest % id for the best taxonomic hit in each sequence (in some, but not all cases, this is 100%)   
      mutate(diff = top_perc - seq_percID) %>% # calculate the difference between the % identity of the top hit and each subsequent taxonomic hit
        arrange(diff)

seq_id_diff %>%
  filter(diff > 0)
```

Now I have the single best entry for each species for each sequence ranked and with the difference between the first and second ranked entries calculated.

For sequences with multiple top hits, where the difference between ranked taxa = 0, I will end up defaulting to genus- or family-level ID. I will do the same for any sequences where the difference betweeen the first and second ranked taxa is < 2%.


Figure out which differences are > 2% and eliminate those first?
```{r}
# filter out any taxa that are >2% less matching identity than the top taxonomic hit for a given sequence
to_remove_low_perc_hits <- seq_id_diff %>%
  ungroup() %>%
  group_by(qseqid) %>%
  filter(diff > 2)

keepers <- seq_id_diff %>%
  anti_join(to_remove_low_perc_hits)
```


```{r}
# this data frame includes only those taxonomic hits that should be considered.
# so now I need to determine whether they should be assigned to genus, family, order, etc. 
singletons <- keepers %>%
  select(qseqid) %>%
  tally() %>%
  filter(n == 1)

# these are the seqs that now have only a single match
singleton_df <- singletons %>%
  left_join(keepers) %>%
  select(-n) %>%
  bind_rows(single_spp_seqs) %>% # combine the single spp data
  mutate(taxonomic_level = "species") %>%
  mutate(taxon = species)

## Genus-level matches
# remove the singletons from the bigger df 
single_genus <- keepers %>%
  anti_join(singleton_df)%>% # at best, these should be genus-level matches
  group_by(qseqid, genus) %>%
  tally() %>%
  ungroup() %>%
  group_by(qseqid) %>%
  tally() %>%
  filter(n == 1) %>% # seqs that match a single genus
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "genus") %>%
  mutate(taxon = genus)
```

okay so i have species and genus level ID's in this data... 

Modify the singleton_df to include the right variable headers
```{r}
single_spp <- singleton_df %>%
  select(-perc_id, -length, -sscinames, -n_taxa) %>%
  mutate(taxonomic_level = "species") %>%
  mutate(taxon = species)
```

```{r}
# recombine the full data set now that the appropriate level of taxonomy has been determined
sorted_tax_df <- bind_rows(single_genus, single_spp)
```

Create output taxonomy data frames
```{r}
uncollapsed_taxonomy <- sorted_tax_df %>%
  select(-top_perc, -id_rank) %>%
  unique()

# and then collapse that down to just a single taxon per ASV
collapsed_taxonomy <- uncollapsed_taxonomy %>%
  select(qseqid, taxon, taxonomic_level) %>%
  unique() %>%
  rename(ASV = qseqid)
```


```{r}
write.csv(collapsed_taxonomy, "/genetics/edna/workdir/crabs/20230216/trimmed/filtered/outputs/asv_taxonomy_blastn.csv")
```



